---
title: "TEOS Hydrostation S Data Lab"
author: "Mileisha L. Velázquez"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
### Load required libraries
```{r,  include=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(gsw)
```
## Import Data: 

Names of columns are separated by commas and white spaces, meanwhile the observations within the rows are separated by only white spaces. This requires rearranging data so that the we may view both the data with the correct column names 


```{r, include=TRUE, echo=TRUE, warning=FALSE, results='asis', message= FALSE }

hydrostation_bottle <- read_delim("hydrostation_bottle.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE, skip = 31)


hydrostation_bottle_names <- read_csv("hydrostation_bottle.txt", 
    skip = 30)

colnames(hydrostation_bottle)=colnames(hydrostation_bottle_names)

view(hydrostation_bottle)
```

# Variable Names and Units: 
```
yyyymmdd = Year Month Day   
decy   = Decimal Year     
time   = Time (hhmm)      
latN   = Latitude (Deg N) 
lonW   = Longitude (Deg W)
Depth  = Depth (m)                  
Temp   = Temperature ITS-90 (C) 
Pres   = CTD Pressure (dbar)   
CTD_S  = CTD Salinity (PSS-78)      
Sal1   = Salinity-1 (PSS-78)        
Sig-th = Sigma-Theta (kg/m^3)       
O2(1)  = Oxygen-1 (umol/kg)          
OxFixT = Oxygen Fix Temp (C)        
Anom1  = Oxy Anomaly-1 (umol/kg)    
// Quality flags //
-999  = No data
 0 = Less than detection limit
```

## Plots: 
```{r, fig.cap= "Figure 1. Clear season signal is observed for sigma-theta" }

hydrostation_bottle %>% 
  filter(`Sig-th` !=-999 & Depth <20) %>% #filter out -999, no data flag and depth for better visual
  ggplot()+geom_line(aes(x=decy,y=`Sig-th`)) #use to visualize patter in data 
```
```{r, fig.cap=" Figure 2. Relationship between temperature and density. Both are strongly correlated"}
hydrostation_bottle %>% 
  filter(`Sig-th` !=-999 & Depth <20) %>% 
  #filter out -999, no data flag and depth for better visual
  ggplot()+geom_point(aes(x=Temp,y=`Sig-th`))#there are two outliers
```

### What to consider 
- Density data from 1988-present, but salinity data from 1950s-present 

- Calculate seawater density form 1950s - present

- TEOS-10 usage for this procedure 


## TEOS-10 Toolbox in Packag seacarb 
```{r}
#gsw
#gsw_sigma0
 #requires SA (absolute salinity) and CT (conservative temperature)

#gsw_SA_from_SP
#sea pressure in (dbar), log, lat, practical sal
```


```{r}
#Pressure plot missing points
hydrostation_bottle %>% 
  ggplot()+
  geom_point(aes(x=decy, y=Pres))
```



```{r}
#Depth data for all time series data 
hydrostation_bottle %>% 
  ggplot()+
  geom_point(aes(x=decy, y=Depth))
```

## Calculate Sea Pressure from Depth Using GSW Function

```{r, message=FALSE, warning=FALSE}
hydrostation_bottle=
  hydrostation_bottle %>% 
mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN))

#Check new variable in plot
hydrostation_bottle %>% 
  ggplot()+
#strong 1:1 agreement between measured pressure and calculated pressure
  geom_point(aes(x=Pres, y=Pres_gsw))
```

## Calculate Absolue Salinity from Practical Salinity
```{r, warning=FALSE}
hydrostation_bottle %>% 
  ggplot()+
  geom_point(aes(x=decy, y=Sal1))

hydrostation_bottle=
  hydrostation_bottle %>% 
mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>% 
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw, 360-lonW, latN))

#Check
hydrostation_bottle %>% 
  ggplot()+
  geom_point(aes(x=decy, y=S_abs_gsw))

#Another way to check 
hydrostation_bottle %>% 
  filter(Sal1!= -999) %>% 
  ggplot()+
  geom_point(aes(x=Sal1, y=S_abs_gsw))
```

## Calculate Conservative Temperature 

```{r}
#gsw_CT_from_t
#absolute salinity , in situ temp (ITS-90), & sea pressure 

#Add line to calculate CT 
hydrostation_bottle=
  hydrostation_bottle %>% 
  filter(Sal1!=-999) %>%
mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>% 
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw, 360-lonW, latN)) %>% 
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw, Temp, Pres_gsw))
   
#Check data
  hydrostation_bottle %>%
  filter(Sal1!=-999) %>% 
  ggplot()+
  geom_point(aes(x=Temp, y=T_cons_gsw))
  
  
```

```{r}
#Still missing data 
  hydrostation_bottle=
  hydrostation_bottle %>% 
  filter(Sal1!=-999) %>%
    filter(Temp!=-999) %>% 
mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>% 
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw, 360-lonW, latN)) %>% 
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw, Temp, Pres_gsw))
  
  #Check again
  hydrostation_bottle %>%
  filter(Sal1!=-999) %>%
   filter(Temp!=-999) %>% 
  ggplot()+
  geom_point(aes(x=Temp, y=T_cons_gsw))
```


## Finally Calculate Sigma-theta: requires SA (absolute salinity) and CT (conservative temperature)
```{r}
#FOR HW 
  HydroS=
  hydrostation_bottle=
  hydrostation_bottle %>% 
  filter(Sal1!=-999) %>%
    filter(Temp!=-999) %>% 
  #calculate pressure
mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>% 
  #calc. absolute salinity
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw, 360-lonW, latN)) %>% 
  #calc.conservative temperature
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw, Temp, Pres_gsw)) %>% 
  #calc. sigma-theta
  mutate(Sig_th_gsw=gsw_sigma0(S_abs_gsw,T_cons_gsw))

HydroS %>% 
  filter(`Sig-th`!=-999) %>% 
  ggplot()+
  geom_point(aes(x=`Sig-th`, y=Sig_th_gsw))
```


### Identify and Fix Error: 

```{r}

#FOR HW 

#Low sig-th-gsw point
HydroS %>% 
  filter(Sig_th_gsw<0) %>% 
  View()#Sal1 too low, caused low sig-th-gsw calculation 

#Calculate Sig-th-gsw with CTD_S instead of Sal1 for outlier point 
HydroS_correctedS_a=  
HydroS %>% 
filter(Sig_th_gsw<0) %>% 
  mutate(S_abs_gsw=gsw_SA_from_SP(CTD_S,Pres_gsw, 360-lonW, latN)) %>% 
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw, Temp, Pres_gsw)) %>% 
  mutate(Sig_th_gsw=gsw_sigma0(S_abs_gsw,T_cons_gsw))
  
HydroS_correctedS_b=  
HydroS %>% 
filter(Sig_th_gsw>0)

HydroS_corrected = rbind(HydroS_correctedS_a, HydroS_correctedS_b)
HydroS_corrected %>% 
  filter(`Sig-th`!=-999) %>% 
  ggplot()+
  geom_point(aes(x=`Sig-th`, y=Sig_th_gsw))
```


# HW: Fix binded data section - add to previous version of HydroS variable 
instead of having two data frames and combining them 
